Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-07 Dail_Mad_ext.tex
intro and overview
intro

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance---hypotheses relating to the importance and existence of
phenomenon such as density-dependent population regulation, population
cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.
%Failure to account for process
%variation and observation error can bias the estimators of abundance
%and related parameters.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Conventional state-space
models are simply time-series models in which the true state of the
system (e.g. population size during each year) is observed
imperfectly. One reason for the widespread adoption of state-space
models in ecology is that failure to account for process variation and
observation error can bias estimators of abundance and population
growth parameters \citep{kery_etal:2009}. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots) and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under exponential population growth, the model could be
written as:
%\begin{gather}
%N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
%X_t = N_t + \zeta_t
%\label{eq:ss1}
%\end{gather}
\begin{subequations}
  \label{eq:ss1}
  \begin{gather}
    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
    N_t \sim \text{N}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
    X_t \sim \text{N}(N_t, \zeta_t)
  \end{gather}
\end{subequations}
where $r$ is the intrinsic rate of increase. Although the exponential
growth model shown here is often replaced with density dependent
models, most state-space models share the common feature that the
error terms are assumed to be i.i.d random effects distrubted as
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Specifically, the Markovian
nature of the model implies the following likelihood
\begin{equation}
\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
[N_t|N_{t-1},\Theta]
\label{eq:L1}
\end{equation}
The first term is the probability density for the first observation in
the time series, $X_1$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation.
%This problem has resulted in several papers with
%interesting names like ``multi-modal likelihoods and ridges etc...''
Two options have been proposed handling the identifiability
problem. First, one can assume that the first observation in the time
series is made without error. Second, one can assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} described methods for fitting
models with non-Gaussian distributions for the process and observation
errors. Observation models with more intuitive interpretations have
been proposed by \citet{kery_etal:2009}. \citet{lele_etal:1998} developed
models allowing for inference about spatial and temporal variation in
abundance, and their developments also resolved the problems of
non-identifiability for the parameters of the initial state at time
$t=1$. In spite of these improvements, until recently, few methods
existed to address each issue simultaneously, and even fewer of these
methods have been made readily available to applied ecologists.

In this paper, we focus on the model of
\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models. Specifically, the model (1) recognizes
the discrete nature of count data and the underlying true abundance
state, (2) includes a realistic observation model with explicit
detection probability parameters (3) has estimable parameters,
including those describing the initial state at time $t=1$, and (4)
allows for inference about both temporal and spatial variation in
abundance. In some respects, the DM model can be viewed as a specific
manifestation of the general state-space model formulated by
\citep{buckland_etal:2004}. However, their model was designed to allow
for inference about demographic parameters, which will typically
require mark-recapture data. Although this may be viewed as the
gold standard in many studies of wildlife populations, in this paper,
we focus on the analysis of count data, which is the typical source of
information used in state-state modeling applications, and one of the
most widely available forms of ecological data---it is also the data
required by the DM model.

@@

Although partitioning population growth into survival and recruitment
is informative, it is not needed for all applications.  Furthermore, a
simpler model would have several merits: faster running time, fewer
total model combinations, and possibly more realistic estimates.
Therefore, we developed a version of the DM model that estimates and
models population growth directly.  In this model, Eq.\ref{eq:Nt} is
simplified to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
\label{eq:exp}
\end{equation}
where $r$ represents the instantaneous population growth rate.  This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This model is based
on the \citet{ricker:1954} discrete time version of the logistic population
growth model.  We also implemented the \citet{gompertz:1825} density-dependent
model:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1})/\log(K)))
\label{eq:gomp}
\end{equation}
where the log of $N_{it-1}$ is only taken when $N_{it-1}>0$
(when $N_{it-1} \equiv 0$, the full expression simplifies to 0 anyway).  Here
the interpretations of $r$ and $K$ are similar to in the Ricker model, but
$K$ is constrained to be greater than one.

>>>>

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.
%Failure to account for process
%variation and observation error can bias the estimators of abundance
%and related parameters.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time-series models in which the true state of the
system (e.g. population size during each year) is observed
imperfectly. One reason for the widespread adoption of state-space
models in ecology is that failure to account for process variation and
observation error can bias estimators of abundance and population
growth parameters.
% \citep{kery_etal:2009}.
For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under exponential population growth, the model could be
written as:
%\begin{gather}
%N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
%X_t = N_t + \zeta_t
%\label{eq:ss1}
%\end{gather}
\begin{subequations}
  \label{eq:ss1}
  \begin{gather}
    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
    N_t \sim \text{N}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
    X_t \sim \text{N}(N_t, \zeta_t)
  \end{gather}
\end{subequations}
where $r$ is the intrinsic rate of increase. Although the exponential
growth model shown here is often replaced with density dependent
models, most state-space models share the common feature that the
error terms are assumed to be i.i.d random effects distrubted as
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Specifically, the Markovian
nature of the model implies the following likelihood
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    [N_1&] \\ \label{eq:ss1b}
    [N_t&|N_{t-1}] &\text{for} \; t=2,\hdots,T \\ \label{eq:ss1c}
    [X_t&|N_t]  &\text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
%\begin{equation}
%\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
%[N_t|N_{t-1},\Theta]
%\label{eq:L1}
%\end{equation}
The first term is the probability density for the first observation in
the time series, $X_1$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation.
%This problem has resulted in several papers with
%interesting names like ``multi-modal likelihoods and ridges etc...''
Two options have been proposed handling the identifiability
problem. First, one can assume that the first observation in the time
series is made without error. Second, one can assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citep{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their models did not include any serial
depedence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. To do so, recent work has sought to combine multiple
sources of information, such as counts and mark-recapture data, to
allow for inferences about population processes such as mortality and
survival, rather than just population growth \citep{besbeas_etal:2002,
  buckland_etal:2004, schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially when inferences are required
at large spatial scales. Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

In this paper, we focus on the model of
\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data. The model (1) recognizes
the discrete nature of count data and the underlying true abundance
state, (2) includes a realistic observation model with explicit
detection probability parameters (3) has estimable parameters,
including those describing the initial state at time $t=1$, and (4)
allows for inference about both temporal and spatial variation in
abundance.

\begin{comment}
  In some respects, the DM model can be viewed as a specific
  manifestation of the general state-space model formulated by
  \citep{buckland_etal:2004}. However, their model was designed to
  allow for inference about demographic parameters, which will
  typically require mark-recapture data. Although this may be viewed
  as the gold standard in many studies of wildlife populations, in
  this paper, we focus on the analysis of count data, which is the
  typical source of information used in state-state modeling
  applications, and one of the most widely available forms of
  ecological data---it is also the data required by the DM model.
\end{comment}

@@

Although partitioning population growth into survival and recruitment
is informative, it is not needed for all applications.  Furthermore, a
simpler model would have several merits: faster running time, fewer
total model combinations, and possibly more realistic estimates.
Therefore, we developed a version of the DM model that estimates and
models population growth directly.  In this model, Eq.~\ref{eq:Nt} is
simplified to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
\label{eq:exp}
\end{equation}
where $r$ represents the instantaneous population growth rate.  This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This model is based
on the \citet{ricker:1954} discrete time version of the logistic population
growth model.  We also implemented the \citet{gompertz:1825} density-dependent
model:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1})/\log(K)))
\label{eq:gomp}
\end{equation}
where the log of $N_{it-1}$ is only taken when $N_{it-1}>0$
(when $N_{it-1} \equiv 0$, the full expression simplifies to 0 anyway).  Here
the interpretations of $r$ and $K$ are similar to in the Ricker model, but
$K$ is constrained to be greater than one.
